Experiences Gamifying Developer Practices

### 2013-10-20 paper_id1.tex
paper submitted to everyone for review on Friday AM

As software development practices evolve, toolsmiths face the continuous challenge of communicating new practices and tools to the community.   We tested an idea with industrial software developers that adding game-like feedback to the development environment focused on improving practices and tool use would improve adoption of tools and practices for code navigation.  We conducted this study in three stages: a pre-study to validate our ideas, a pilot of a tool called Blaze that implements some game features to motivate better navigation practices, and a post-pilot survey conducted via interviews.  We found that a software engineering game with competitive elements influences developers towards changing their practices.
%We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recommendations on ways they can improve.
%
% In addition to training, discussing, and presenting software engineering practices and tools, we demonstrate a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).  
%
%We determine through a survey of 130 developers the things we can do to motivate them to adopt 
%Blaze monitors developer practice and tool use over the long term and reports the lasting change a tool deployment or communication effort makes in everyday habits. 
%
% Embedding Blaze in the developer's work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.   Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.  
% With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.  
%
%Data collected from the pilot group show a dichotomy of navigation and code understanding practices used by developers even the same developer in different situations.  We observed that the duration of an edit session is more correlated with categories of activity that benefit from improved practices and tools. The Blaze tool was well received by most participants, however, navigation practices were not significantly affected by the Blaze feedback mechanism.  We conclude that more proactive interventions are necessary to encourage developers to change their code navigation habits.
\end{abstract}

@@

Sascha is a developer with years of experience maintaining code who knows how to navigate through code fairly well.  Sascha heard about some other navigation tools, but kept using the built-in features of the Visual Studio IDE that served him well.  When Sascha's company setup a game around code navigation in Visual Studio, he became interested in being a part of the fun.  The game gave bonus points for using Sando, a new code search tool, so he started using it instead of some of the built-in search commands.  Sascha soon topped the game's leaderboard and discovered that motivated him to try the next tool to became part of the game.

Like Sascha, software developers who maintain code face additional challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.  Over the years, developers learn techniques to analyze code and locate areas that are relevant to an enhancement.  As new practices and tools get introduced, developers may not learn about them unless they interact with a peer developer who uses them\cite{wbsnipes:Hill2011Peer}.    
%For instance, the commands they use to navigate the code base in the development environment may make them less efficient than using newer practices or tools.  Robillard et al. show that developers who use commands that allow them to follow code structure are more effective and efficient than developers who browse through code \cite{wbsnipes:Robillard2004How}.

Enter the toolsmith,  who responds to developers' challenges by creating tools such as code search tools that make the job easier and more efficient.  The toolsmith needs to spread the word about advanced tools they develop and determine their effect on developer practices over the long term.  Is there something we can do that encourages developers to adopt new practices and tools that also provides feedback to developers and toolsmiths on the resulting improvements?

One option for motivating people to adopt new practices is to apply elements of game mechanics or ``gamify'' the practice. Gamification is the incorporation of game elements into an activity that people do not typically consider a game\cite{2013Oxford}.  We envisioned a tool in our prior work that would motivate developers to adopt new practices or tools by providing constant feedback and allowing them to compare their improvements to how their peers are doing  \cite{Snipes2013Towards}.  This tool called Blaze, creates a competitive game that gives continuous feedback to the developer on adopting the improved practice.  The more they change their practices to adopt new techniques or tools, the higher they score in the game.

Would software developers writing serious software be motivated by this gamification concept? To answer this question, this work contributes an assessment of developers' response to game-like elements tied to their use of new tools and practices.  We assess their response to the idea of gamification and relevant concerns through a survey of 130 software developers.  Then we conduct a pilot of a tool, Blaze, that introduces game elements into the developer's IDE focused on improving navigation practices.  With the pilot feedback we determine the acceptance, influence, and effect of deploying the Blaze tool in the pilot group.

The rest of this paper is organized as follows:  Section 2 reviews Related work, Section 3 reviews results from the pre-study survey, Section 4 Discusses the design considerations for developing the Blaze tool/game, Section 5 Discusses the design of the Blaze pilot study and participant selection,  Section 6 results from the pilot study and post-pilot survey, Section 7 Threats to Validity, and Section 8 has our conclusions.
%This is precisely what we set out to study for this work. 
%
%Our first step was to challenge this question through a pre-study survey that queried both whether developers would be interested in gamifying their workspace and what aspects of that space would be most interesting to them.
%
%After confirming a good level of interest from the pre-study survey, we identified a set of practices and defined a game meeting expectations from the developers feedback.  The game, implemented in a Visual Studio add-in called Blaze, integrates into the Visual Studio workspace and establishes an anonymous competition between developers to improve their navigation practices. 
%
%During the pilot of Blaze we observed the variation in developer navigation practices and identified some potential reasons for this variation.  We conducted post-pilot survey and feedback sessions that evaluated developers' acceptance of game elements in the tool and their desires for additional functionality.

\section{Related Work}
%update and expand
\subsection{Monitoring Practice Studies }
Johnson and Kou defined Zorro\cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.  Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.  Using the distinct events that developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario.  In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy in classifying episodes into their proper TDD scenarios.  The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.  In our study, we attempt this leap to present data collected directly to developers attempting to influence them to change their practices.

Murphy-Hill, Parnin, and Black \cite{V:MurphyHill2012How} use the Mylyn Monitor to explore whether or not developers use the automated refactoring tools present in Eclipse.  They look for specific refactoring commands in Eclipse and determine the amount of time developers use tools versus hand refactoring the code.  For this study we build a tool that captures the use of all commands within the Visual Studio environment including refactoring, navigation, and edit commands.

Robillard, Coelho and Murphy explore hypotheses around how developers can be more effective at performing a maintenance task \cite{wbsnipes:Robillard2004How}.  Key conclusions are that developers are more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.  This builds on this work by testing methods for increasing the use of structured navigation in developers' everyday practice.

The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.  Studies conducted using PROM include a series of studies by it's creators on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.  The PROM tool and studies that use it track the time spent editing files and methods from the IDE and code metrics obtained trough source code analysis.  Applying these data to specific research questions such as whether refactoring improves productivity \cite{Moser2008Case} shows the utility of combining automated effort data with code metric data.  Taking a different focus on techniques, our study used more detailed events captured from the IDE  to detect how the user was navigating through the code in addition to knowing how long the editor is open for a particular file.

Murphy-Hill et al. study a large usage history data set and apply several different algorithms to accurately suggest to users new commands in their environment to use.\cite{MurphyHill2012Improving} Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history. Novice users were better served by recommendation algorithms that did not require a long history, where more expert users benefited from more sophisticated algorithms that included more history. The ``Most Widely Used'' algorithm that recommends commands based on the collective usage profile of the team performed nearly as well as the more sophisticated algorithms. The paper presents the methodology for making recommendations and the evaluation of the accuracy of the recommendations. Our work goes the next step of evaluating whether a specific recommended practice gets adopted by developers when influenced through gamification methods.

@@

When discussing gamification, folks raise a concern about replacing an activity that is intrinsically motivated with extrinsic motivation provided by points and achievements.  To discover what is behind this concern, several studies are useful to understand what should be done.

Early studies attempt to evaluate achievement motivation in students and relate that to their performance in school.  The study by Hermans describes results from applying questionnaire to evaluate achievement motivation in students \cite{wbsnipes:Hermans1970Questionnaire}.  Each question is ranked for its correlation to achievement motivation. Some key questions about the diligence with which students approach their work are correlated with achievement motivation.

Maehr proposes an affirmative theory whereby  individuals achieve as a member of a social group choosing the behavior that will meet the expectations and values of the group that is significant to them \cite{wbsnipes:MaehrCulture}.  The presence of a social group where the person is motivated to belong and excel within is important for creating achievement motivation.  Maehr states that since achievement behaviors can be triggered by circumstance, that they are present in most people regardless of background.

Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering \cite{wbsnipes:Beecham2008Motivation}.  They provide a useful inventory of aspects of the job motivate software engineers with the most common being ''the work'' motivates us. The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making. The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills/tackle challenges. 
Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository \cite{Singer2012It}.  The communication mechanism enabled students to see each others' progress and resulted in more frequent commits than baseline.  Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.  The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.  Results show an increase in the frequency of commits at a statistically significant level.  This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.

Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development \cite{Dubois2013Understanding}. Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.  They provide guidance on assessing gamification in the software development domain that includes analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.  Dbois and Tamburrelli demonstrate this guidance using a study of students working on a class project.  Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.  The subject group can see each others' rankings while the control group only sees their personal rankings.  Results though inconclusive indicate that competition increases the desired results.  In our study we apply both individual feedback and competitive based feedback (anonymized) to participants and seek to observe a change in their behavior after an intervention.

@@

Understanding the concerns with maintaining developers'  intrinsic motivation while adding elements of achievement motivation, we conducted an assessment survey questionnaire to determine whether the ABB developer community would be receptive to a gamification approach to software development.   Another question asked about ways of motivating developers to use training, practices, and tools.   We checked on who they would be comfortable sharing data from their development activities with.   Finally we asked what kind of awards would be most welcome such as badges and points. The developer survey received over 130 responses from software developers at ABB.

@@

Using a question from the Saatchi and Saatchi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB ``How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace''?  We segmented results by country in Figure \ref{fig:gamification} and found that gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.  The responses indicating that developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.  For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.  Because we nearly identically reused their original question, we compare the overall response from our survey to the Saatchi and Saatchi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based.  Our results land on both sides of the 75\% of 18-45 year olds that Saatchi and Saatchi found were at least somewhat interested.

@@

We asked developers how likely they would be to try tools and practices recommended to them through an automated usage tracking system.  Answers showed that 95\% of the developers surveyed are likely to try the recommended tools and practices.  This indicates that a recommendation system would positively impact the deployment of software engineering tools and practices at ABB.

Another question was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.  The question provided a graduated scale for the scope of who data were shared with and divided the questions between whether that sharing was anonymous or not. Figure \ref{fig:comfortwithsharing}  shows that over 90\% of respondents are comfortable with sharing either anonymous or non-anonymous information with their team members. 
Their comfort level decreases as the scope of who the information is shared with increases.  People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data were not kept anonymous.  
The first conclusion from these questions on sharing data is that people are very willing to share information that could help their team. Second, sharing within the company is acceptable if we take care to make the data anonymous.

@@

We asked developers what would motivate them to use tools and practices and gave them 4 choices to rank.  The top motivator for using tools and practices is coworker recommendation with 75\% of respondents in Figure \ref{fig:toolandpracticemotivators}   ranking in 1st or 2nd place. The lowest was having badges posted on their social network profile where 33\% ranked this in 1st or 2nd place.   Badges also ranked low in motivations for taking more software engineering training.  We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.  Since a competition among teams ranks slightly higher than management mandate, we have the choice to apply one or the other or both as motivation for using key practices.

@@

Developers also ranked the desirability of awards for different actions.  Overall 68\% of developers ranked receiving an award for consistently performing quality practices as either 1st or 2nd place.  This supports a conclusion that focusing on awarding good practices would motivate developers to do them consistently.

 To summarize, the high-level conclusions from each key question are:
\begin{itemize}[itemsep=0mm]
\item 95\% of developers responding would try tools and practices suggested by an automated recommendation system 
\item Developers favor a team-oriented competition over individual awards
\item Awards for performing good quality development practices ranked highest
\item Over 90\% of respondents approved of sharing data with their team members
\end{itemize}
Thus, this pre-study survey provided a strong positive indicator for going forward with a software engineering tool that focused on sharing developer data and rewarding developers for using good practices and tools.

@@

Following the pre-study findings, we implemented a game-like tool for developers to use in Visual Studio.  The tool focused on the objectives of motivating the adoption of good practices by including elements of games in the development environment.  The following sections describe how we selected a practice suitable for gamification, designed the game, and included competitive elements in the game.

@@

We defined the first criteria for selecting a practice based on prior work on applying gamification to other domains outside software development. 
In their paper on game design patterns, Hamari et al. issue a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework} .  Thus we avoided selecting practices like bug fixing or task completion because they are requirements that must be done.

Another criteria is imposed by the instrument scope where we must limit ourselves to practices that exist entirely in the IDE that can be monitored using Blaze.  We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and frequent CM submissions.

@@

Studies of developer effectiveness by Robillard et al.\cite{wbsnipes:Robillard2004How} identify structured navigation as a practice that more effective developers use when maintaining programs.   Structured navigation fits the requirements of a practice that is contained in the IDE, has proven benefits, lends itself to measurement and scoring, is not an assigned task, and leaves room for developers to excel.  Thus we selected structured navigation as the practice to demonstrate gamification.

@@

To setup the game, the Blaze tool provides an XML configuration file where the researcher can configure Blaze to categorize and assign points to the commands that are part of a software engineering practice.  Blaze allows the researcher to define multiple command category levels.  Thus we can categorize commands that are navigation then further classify them into structured and unstructured navigation.

We categorize commands as structured navigation when they allow the developer to locate a code element by following the program structure from another element.  These commands include Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R} that are built-in to Visual Studio and follow the structured graph of the code.

Commands that we considered as unstructured navigation include scrolling through a file using cursor keys or the scroll-bar, opening files from the Solution Explorer, Find commands (e.g. Find in Files), children of these commands (e.g. Find Next), and Go-to line number.

@@

Gamification practices consider the feedback of achievements and points as critical to a successful outcome.  The design of achievement awards should gradually encourage the participant to reach higher and higher levels of success in the activity.  Achievements must feel earned to the participant so that internally they recognize the effort to get the award was significant and feel satisfied  \cite{wbsnipes:Hamari2011Framework}.  Hamari and Eranti describe a general relationship between achievements in games and regular game play      In games achievements are typically a parallel scoring system to the main game play.  They make the game more engaging providing multiple ways to increase your score and multiple challenges in one interface.

Using these guidelines we assigned points to structured navigation commands which received 1 point each while unstructured navigation commands received 0 points.    We gave additional emphasis to using the Sando search tool by assigning it 10 points.  To create levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we applied an exponential curve based on points scored.  Users initially level up after a working day or two extending to achieving the next level after a week or even a month following an exponential curve.  We designed the points and levels such that above average developers can complete all levels during the study period.

@@

By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition among teammates.    The leaderboard allows developers to see their own score in relation to the top 5 people on the team.  However, the identifier of the other participants was generated as 3 capital letters assigned based on their position on the leaderboard.  Not knowing who of your colleagues was ahead of you could be a factor that reduced motivation but avoids one potential source of conflict in an intact team.

@@

It was important to gain the confidence of the participants that their data were kept confidential so their activities would be close as possible to real-world. if the participants felt like they are being tracked and the data is going to use for some study that would identify them to their peers or managers, then they may deviate from the natural development style.

>>>>

As software development practices evolve, toolsmiths face the continuous challenge of communicating new practices and tools to the community.   We tested an idea with industrial software developers that adding game-like feedback to the development environment would improve adoption of tools and practices for code navigation.  We conducted this study in three stages: a pre-study to validate our ideas, pilot of a tool with some game features to motivate better navigation practices, and feedback interviews for a post-pilot survey.  We found that a software engineering game with competitive elements influences some developers towards changing their practices.
%We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recommendations on ways they can improve.
%
% In addition to training, discussing, and presenting software engineering practices and tools, we demonstrate a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).  
%
%We determine through a survey of 130 developers the things we can do to motivate them to adopt 
%Blaze monitors developer practice and tool use over the long term and reports the lasting change a tool deployment or communication effort makes in everyday habits. 
%
% Embedding Blaze in the developer's work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.   Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.  
% With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.  
%
%Data collected from the pilot group show a dichotomy of navigation and code understanding practices used by developers even the same developer in different situations.  We observed that the duration of an edit session is more correlated with categories of activity that benefit from improved practices and tools. The Blaze tool was well received by most participants, however, navigation practices were not significantly affected by the Blaze feedback mechanism.  We conclude that more proactive interventions are necessary to encourage developers to change their code navigation habits.
\end{abstract}

@@

Sascha is a developer with years of experience maintaining code who knows how to navigate through code fairly well.  Sascha heard about some other navigation tools, but kept using the built-in features of the Visual Studio IDE.  When Sascha's company started a game around code navigation in Visual Studio, he became interested in being a part of the fun.  The game gave bonus points for using Sando, a new code search tool, so he started using it instead of some of the built-in search commands.  Sascha soon topped the game's leaderboard and discovered he became motivated to try the next practices included in the game.

Like Sascha, software developers who maintain code face additional challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.  Over the years, developers learn techniques to analyze code and locate relevant areas for an enhancement.  As new practices and tools get introduced, developers may not learn about them unless they interact with a peer developer who uses them\cite{wbsnipes:Hill2011Peer}.    
%For instance, the commands they use to navigate the code base in the development environment may make them less efficient than using newer practices or tools.  Robillard et al. show that developers who use commands that allow them to follow code structure are more effective and efficient than developers who browse through code \cite{wbsnipes:Robillard2004How}.

Enter the toolsmith,  who responds to developers' challenges by creating tools such as code search tools that make the job easier and more efficient.  The toolsmith needs to spread the word about advanced tools they develop and determine their effect on developer practices over the long term.  Is there something we can do to encourage developers to adopt new practices and tools that also provides feedback to developers and toolsmiths on the resulting improvements?

One option for motivating people to adopt new practices is to apply elements of game mechanics or ``gamify'' the practice. Gamification is the incorporation of game elements into an activity that people do not typically consider a game\cite{2013Oxford}.  We envisioned a tool in our prior work to motivate developers by providing constant feedback and allowing them to compare their improvements to how their peers are doing  \cite{Snipes2013Towards}.  This tool called Blaze, creates a competitive game with continuous feedback on how well the developer adopts the improved practice.  The more they change their practices to adopt new techniques or tools, the higher they score in the game.

Would software developers writing serious software be motivated by this gamification concept? To answer this question, this work contributes an assessment of developers' response to game-like elements tied to their use of new tools and practices.  We assess 130 developers' response and relevant concerns about the idea of gamification through a web-based survey.  We piloted Blaze, our tool that introduces game elements focused on improving navigation practices into the developer's IDE.  With the pilot feedback we determine the acceptance, influence, and effect of deploying the Blaze tool in the pilot group.

The rest of this paper is organized as follows:  Section 2 reviews related work, Section 3 reviews results from the pre-study survey, Section 4 discusses the design considerations for developing the Blaze tool/game, Section 5 discusses the design of the Blaze pilot study,  Section 6 results from the pilot study and post-pilot survey, Section 7 covers threats to talidity, and Section 8 has our conclusions.
%This is precisely what we set out to study for this work. 
%
%Our first step was to challenge this question through a pre-study survey that queried both whether developers would be interested in gamifying their workspace and what aspects of that space would be most interesting to them.
%
%After confirming a good level of interest from the pre-study survey, we identified a set of practices and defined a game meeting expectations from the developers feedback.  The game, implemented in a Visual Studio add-in called Blaze, integrates into the Visual Studio workspace and establishes an anonymous competition between developers to improve their navigation practices. 
%
%During the pilot of Blaze we observed the variation in developer navigation practices and identified some potential reasons for this variation.  We conducted post-pilot survey and feedback sessions that evaluated developers' acceptance of game elements in the tool and their desires for additional functionality.

\section{Related Work}

The two areas of related work we discuss are research studies using other monitoring methods and studies applying motivation and gamification to other areas.

\subsection{Monitoring Practice Studies }

Robillard, Coelho and Murphy explore hypotheses around how developers can be more effective at performing a maintenance task \cite{wbsnipes:Robillard2004How}.  Robillard et al. say developers are more successful finding and fixing bugs when they create a detailed plan for implementing a change, use structured navigation with keyword search or cross-reference search, and only review methods once during their search.  We build on their work by testing methods for increasing the use of structured navigation in developers' everyday practice.

Johnson and Kou defined Zorro\cite{V:Johnson2007Automated}, a system for detecting whether developers use Test Driven Development techniques based on data from Hackystat.  Zorro divides development activities into episodes delimited by events such as configuration management code check-in, start of a unit test run, or start of a build.  Using the distinct events developers follow within these episodes, Zorro determines whether the episode followed Test-Driven Development practices per their classic definition of test first - code - test pass, or a different scenario.  In two student-based studies comparing Zorro classifications with a simultaneous observational screen video, Zorro achieved between 70\% \cite{Kou2010Operational} and 89\% \cite{V:Johnson2007Automated} accuracy when classifying episodes into their proper TDD scenarios.  The studies did not, however, attempt to use or evaluate the influence of instant feedback aimed at encouraging participants to follow the classic definition of Test Driven Development.  In our study, we attempt to present data collected directly to developers in order to influence them to change their practices.

Murphy-Hill, Parnin, and Black \cite{V:MurphyHill2012How} use the Mylyn Monitor to explore whether or not developers use the automated refactoring tools present in Eclipse.  They look for specific refactoring commands in Eclipse and determine the amount of time developers use tools versus hand refactoring the code.  For this study we focus on code navigation as a high-value area for improving developer efficiency.

The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.  Studies conducted using PROM include a series of studies on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.  Studies using the PROM tool correlate the time spent editing with code metrics obtained trough source code analysis.  Applying these data to specific research questions such as whether refactoring improves productivity \cite{Moser2008Case} shows the utility of combining automated effort data with code metric data.  Taking a different focus on techniques, our study used more detailed events captured from the IDE  to detect how the user was navigating through the code in addition to knowing how long the editor is open for a particular file.

Murphy-Hill et al. study a large usage history data set and apply several different algorithms to accurately suggest IDE commands to novice users.\cite{MurphyHill2012Improving} Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history. Novice users were better served by recommendation algorithms that did not require a long history, where more expert users benefited from more sophisticated algorithms that included more history. The ``Most Widely Used'' algorithm recommends commands based on the collective usage profile of the team and performed nearly as well as the more sophisticated algorithms. The paper presents the methodology for making recommendations and evaluate their accuracy. Our work goes takes a further step evaluating how well tools can influence developers to use a specific recommended practice.

@@

When discussing gamification, folks raise a concern about replacing an activity like software development, which is intrinsically motivated, with extrinsic motivation provided by points and achievements.  To discover what is behind this concern, several studies are useful to understand what should be done.

Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering \cite{Beecham2008Motivation}.  They claim most common motivator found in cited references is ''the work'' motivates us. The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making. The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills and tackle new challenges.

Maehr proposes an affirmative theory whereby  individuals achieve as a member of a social group choosing the behavior that meets the expectations and values of the group that is significant to them\cite{wbsnipes:MaehrCulture}.  A key factor in raising achievement motivation is establishing a social group where the person is motivated to belong and excel.  Maehr states that since achievement behaviors can be triggered by circumstance, they are present in most people regardless of background.

Hamari et al. analyze achievements patterns in game systems and provide guidance on defining successful achievement systems.  They recommend  achievement awards be setup as a secondary form of scoring so players have multiple objectives available for a particular game.  Players should feel like they earned achievements in the game, they should not be given for trivial tasks.

Singer and Schneider demonstrate the use of a message board and points for encouraging students to increase their frequency of commits to the source code repository \cite{Singer2012It}.  The communication mechanism enabled students to see each others' progress and resulted in more frequent commits than baseline.  Participants valued the communication and collaboration aspect and some valued the competition enough to change their opinion on optimum commit frequency.  The subsequent thesis by Singer \cite{Singer2013a} describes results of an experiment conducted across two iterations of a class where active feedback on commits was deployed to one course and another course served as the control group where commit frequency was simply monitored.  Results show an increase in the frequency of commits at a statistically significant level.  This study uses fine grained instrumentation to identify practices used between commits such as structured navigation.

Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development \cite{Dubois2013Understanding}. Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.  They provide guidance on assessing gamification in the software development domain including analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.  Dbois and Tamburrelli demonstrate this guidance using a study of students working on a class project.  Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.  The subject group can see each others' rankings while the control group only sees their personal rankings.  Results though inconclusive, show competition increases the desired results.  In our study we provide participants with both individual feedback and anonymized competitive feedback and seek to observe a change in their behavior.

@@

Understanding the concerns with maintaining developers'  intrinsic motivation while adding elements of achievement motivation, we conducted an assessment survey questionnaire to determine whether the ABB developer community would be receptive to a gamification approach to software development.  The developer survey received over 130 responses from software developers at ABB.

@@

Using a question from the Saatchi and Saatchi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB ``How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace''?  We segmented results by country in Figure \ref{fig:gamification} and found gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.  The responses indicate developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.  For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.  Because we nearly identically reused their original question, we compare the overall response from our survey to the Saatchi and Saatchi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based.  Our results land on both sides of the 75\% mark resulting from the Saatchi and Saatchi study.

@@

We asked developers how likely they would be to try tools and practices recommended to them through an automated usage tracking system.  Answers showed 95\% of the developers surveyed are likely to try the recommended tools and practices.  Thus, a recommendation system would positively impact the deployment of software engineering tools and practices at ABB.

Another question was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.  The question provided a graduated scale for the scope of who data were shared with and divided the questions between whether sharing was anonymous or not. Figure \ref{fig:comfortwithsharing}  shows over 90\% of respondents are comfortable with sharing either anonymous or non-anonymous information with their team members. 
Their comfort level decreases as the scope of who the information is shared with increases.  People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data were not kept anonymous.  
The first conclusion from these questions on sharing data is people are very willing to share information that could help their team. Second, sharing within the company is acceptable if we take care to make the data anonymous.

@@

We asked developers what would motivate them to use tools and practices and gave them 4 choices to rank.  The top motivator for using tools and practices is coworker recommendation with 75\% of respondents in Figure \ref{fig:toolandpracticemotivators}   ranking in 1st or 2nd place. The lowest was having badges posted on their social network profile where 33\% ranked this in 1st or 2nd place.   Badges also ranked low in motivations for taking more software engineering training.  We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.

@@

Developers also ranked the desirability of awards for different actions.  Overall 68\% of developers ranked receiving an award for consistently performing quality practices as either 1st or 2nd place.  Thus, focusing awards on good practices would motivate developers to do them consistently.

 To summarize, the high-level conclusions from each key question are:
\begin{itemize}[itemsep=0mm]
\item 95\% of developers responding would try tools and practices suggested by an automated recommendation system 
\item Developers favor a team-oriented competition over individual awards
\item Awards for performing good quality development practices ranked highest
\item Over 90\% of respondents approved of sharing data with their team members
\end{itemize}
Thus, this pre-study survey provided a strong positive indicator for going forward with a software engineering tool focused on sharing developer data and rewarding developers for using good practices and tools.

@@

Following the pre-study findings, we implemented a game-like tool for developers to use in Visual Studio.  The tool focused on motivating the adoption of good practices by including elements of games in the development environment.  The following sections describe how we selected a practice suitable for gamification, designed the game, and included competitive elements in the game.

@@

We defined the first criteria for selecting a practice based on prior work  applying gamification to other domains outside software development. 
In their paper on game design patterns, Hamari et al. issue a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework}.  Thus we avoided selecting practices like bug fixing or task completion because they are simply requirements of the job.

Another criteria is imposed by the instrument scope where we must limit ourselves to practices performed entirely in the IDE so they can be monitored using Blaze.  We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and frequent CM submissions.

@@

Studies of developer effectiveness by Robillard et al.\cite{wbsnipes:Robillard2004How} identify structured navigation as a practice more effective developers use when maintaining programs.   Structured navigation fits the requirements of a practice contained in the IDE, has proven benefits, lends itself to measurement and scoring, is not an assigned task, and leaves room for developers to excel.  Thus we selected structured navigation as the practice to demonstrate gamification.

@@

To setup the game, the Blaze tool provides an XML configuration file where the researcher can configure Blaze to categorize and assign points to the commands that are part of a software engineering practice.  Blaze allows the researcher to define multiple command category levels.  Thus we can categorize commands as navigation then further classify them into structured and unstructured navigation.

We categorize commands as structured navigation when they allow the developer to locate a code element by following the program structure from another element.  These include  built-in Visual Studio commands like Navigate To \textit{Ctrl+,}, Go To Definition \textit{F12}, and Find All References \textit{Ctrl+K,R}  that follow the structured graph of the code.

Commands we considered as unstructured navigation include scrolling through a file using cursor keys or the scroll-bar, opening files from the Solution Explorer, Find commands (e.g. Find in Files), children of these commands (e.g. Find Next), and Go-to line number.

@@

Gamification practices consider the feedback of achievements and points as critical to a successful outcome.  The design of achievement awards should gradually encourage the participant to reach higher and higher levels of success in the activity.  Achievements must feel earned to the participant so they recognize the effort to get the award was significant and feel satisfied  \cite{wbsnipes:Hamari2011Framework}.  Hamari and Eranti describe a general relationship between achievements in games and regular game play      In games achievements are typically a parallel scoring system to the main game play.  They make the game more engaging providing multiple ways to increase your score and multiple challenges in one interface.

Using these guidelines we assigned points to structured navigation commands which received 1 point each while unstructured navigation commands received 0 points.    We gave additional emphasis to using the Sando search tool by assigning it 10 points.  To create levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we applied an exponential curve based on points scored.  Users initially ``level up'' after  working a day or two; the next level may be achieved after a week or even a month following an exponential curve.  We designed the points and levels so above average developers can complete all levels during the study period.

@@

By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition among teammates.    The leaderboard allows developers to see their own score in relation to the top 5 people on the team.  However, the identifier of the other participants was generated as 3 capital letters assigned based on their position on the leaderboard.  Not knowing who of your colleagues was ahead of you could be  reduce motivation, but this avoids one potential source of conflict in an intact team.

@@

It was important to gain the confidence of the participants that their data were kept confidential so their activities would be close as possible to real-world. if the participants felt like they are being tracked and the data would identify them to their peers or managers, then they may refuse to participate or deviate from their normal development style.
